### 2016-01-12 _site/responseToReviewers/mysisReviewerResponse.Rmd

```{r loadLibraries, echo=FALSE, include=FALSE, message=FALSE}
## We start by loading packages and define helper functions
options(digits=3)
set.seed(123)
library(knitr)
# knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(include=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(warning=FALSE)
library(ggplot2)
library(Gmisc)
library(MASS)
library(stringr)
library(mgcv)
library(lme4)
library(ismev)
library(car)
library(betareg)
```

@@

To determine if there was any difference between the two sized nets, we have a few questions of interest. Our main questions of interest are 1) Is there a difference in density of *Mysids* caught between the two nets, 2) Is there a difference in mean length of *Mysids* caught between the two nets, and 3) Does the density when broken down by age and sex class differ between the two net sizes? Note that we will use count and density interchangeable in what follows because we normalize to a one $m^2$ area.

## 1) Is there a difference in density of *Mysids* caught between the two nets?

@@

```{r totalCount, eval=TRUE}
load("~/mysis/data/mysisCountData.RData")                                 ## load count data
```

```{r plotCount, eval=TRUE, include=TRUE}
ggplot(data=mysisCountData, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
  ggtitle("Normalized counts by net size and date")
```

By looking at the distributions of counts across time, we see that they are similar between net sizes, but the normality of the counts is questionable. From the histogram, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August. Because counts cannot be less that zero and often are right skewed, the assumption of normality in the data is questionable. Hence, methods like linear regression, t-tests, and ANOVA that assume normal distributions are not ideal, but, for completeness we begin with a paired t-test.

```{r pariedTTest}
y1 <- mysisCountData$count[mysisCountData$net == "Small Net"]
y2 <- mysisCountData$count[mysisCountData$net == "Large Net"]
pairttest <- t.test(y1, y2, paired=TRUE)
summary(pairttest)
```

@@

```{r totalCountPoisson, message=FALSE}
mns <- by(mysisCountData[, "count"], mysisCountData[, "net"], mean)       ## mean of counts by net size
vars <- by(mysisCountData[, "count"], mysisCountData[, "net"], var)       ## variance of counts by net size
```

We test for a difference in density between the nets formally by fitting a model to compare if the count of *Mysids* caught by the large net is the same as the count of *Mysids* caught by the large net, after normalizing for the area of the two nets and controlling for covariates. The class of models known as generalized linear models (glms) offers a rigorous solution for modeling count data. These models allow for regression models that account for the effects of covariates like in linear regression but do not assume a normal distribution. For count data, there are two natural distributions to use: the Poisson distribution and the negative binomial distribution. The Poisson distribution has the strong assumption that the mean equals the variance, which is often not met in practical datasets. In our data, the count means are $(`r I(mns)`)$ and the count variances are $(`r I(vars)`)$, for the small and large nets, respectively. Therefore, we use a negative binomial model
$$
\label{neg:bin}
y_i \sim \mbox{NegBin}(\mu_i, \phi),
$$
where $\mu_i$ is the mean of the negative binomial distribution for sampling event $i$ and $\phi$ is an overdispersion parameter that allows for the mean and variance to be different. We model $\mu_i$ with covariates using the canonical log link function
$$ 
\mu_i = log(\mathbf{X}_i \boldsymbol{\beta}),
$$
where $\mathbf{X}_i$ is the set of covariates for observation $i$. Then we perform inference on our coefficients $\boldsymbol{\beta}$, where the interpretation of $\beta_j$ is a percent change in count per unit change in $X_{ij}$.

```{r totalCountModel, eval=TRUE}
summary(nbmod <- glm.nb(count ~ net + date + station, data=mysisCountData))
## get coefficient estimates
coefs <- c(nbmod$coefficients, nbmod$theta)           
```

@@

To test for a change in total counts between the two net sizes, we construct a negative binomial regression model that examines the effect of net size while accounting for date of sampling and sampling location. The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes ($z$-value = `r I(round(summary(nbmod)$coef[2, 3], 4))`, $p$-value=`r I(round(summary(nbmod)$coef[2, 4], 4))`).

```{r countModelResults, eval=FALSE}
tableData <- cbind(summary(nbmod)$coeff[1:6, ])
colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
                         "Net Size * August", "Net Size * September")
kable(tableData, digits = 4, format="html")
```

## 2) Is there a difference in mean length of *Mysids* caught between the two nets?
First we examine some histograms and get an idea of what the length data look like.

@@

One could argue that because the larger net catches four times the number of *Mysids*, the large net is more likely to catch *Mysids* at the extremes of the size classes (either really small or really large *Mysids*). To account for this, we perform a paired t-test using a trimmed mean, removing the smallest and largest 5% of the lengths before calculating the sample mean. From the paired t-test using the trimmed means, we find that there is not a significant difference ($t$ = `r pairttestlengthtrim$statistic`, $df$ = `r pairttestlengthtrim$parameter`, $p$-value = `r pairttestlengthtrim$p.value`) in expected counts, although the $p$-value is close to the 0.05 level. As with the untrimmed means, the practical difference is quite small as judged by the histograms and small difference in mean *Mysid* size difference (`r pairttestlengthtrim$estimate` mm).

To better account for the different sample sizes (the group means are from unbalanced sample sizes), we fit a linear model and a robust linear model (that assumes overdispersion in the data). We start by checking for heterogeneity in the group variances.

```{r}
vars <- as.vector(by(mysisLengthData$y, mysisLengthData$label, var))
qplot(vars, main = "var by sampling location", xlab="var")
bart_test <- bartlett.test(mysisLengthData$y, mysisLengthData$label)
levene_test <- leveneTest(mysisLengthData$y, mysisLengthData$label)
```

When looking at the distribution of sampling variance by sampling occasion (10 locations, 2 nets, and 4 different sampling dates gives 80 total sampling occasions), we see that there is some difference in variance of lengths by sampling occasion and Bartlett's test ($p$-value `r format.pval(bart_test$p.value)`) and Levene's test ($p$-value `r format.pval(levene_test$Pr[1])`).

@@

d <- (mean(mysisLengthData$y[mysisLengthData$net == "Large Net"]) - 
        mean(mysisLengthData$y[mysisLengthData$net == "Small Net"])) / SD
```

Although the statistical tests using both the weighted linear model and the robust linear model show statistically significant differences in mean length caught between the net sizes, this is not unexpected because have very large sample sizes and and the regression coefficients are very small relative to the other sources of variation in the model. Of greater interest is whether the observed effect is of practical significance.

The effect size of the mean length difference between the two net sizes is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`). Given a large sample size, a hypothesis test will show statistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05). Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(mysisLengthData$y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest. Another measure of effect size is Cohen's $d$ which measures the difference in means relative to a pooled standard deviation. For our data, Cohen's $d$=`r I(d)`, which implies that the effect of net size is quite small in terms of practical significance.

## 3)  Does the density when broken down by age and sex class differ between the two net sizes?
The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates. A visual inspection of counts by sex class shows no difference in counts by net size, but does show a change in counts over time for each sex class.

### 2016-01-12 responseToReviewers/mysisReviewerResponse.Rmd

---
title: Effect of net size on estimates of abundance, size, age and sex ratio of Mysis
  diluviana Journal of Great Lakes Research - Statistical Analysis
author: "John Tipton"
date: "January 3, 2016"
output: html_document
---

```{r loadLibraries, echo=FALSE, include=FALSE, message=FALSE}
## We start by loading packages and define helper functions
options(digits=3)
set.seed(123)
library(knitr)
# knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(include=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(warning=FALSE)
library(ggplot2)
library(Gmisc)
library(MASS)
library(stringr)
library(mgcv)
library(lme4)
library(ismev)
library(car)
library(betareg)
```
